# Imperfect Competition
Market power implies that at least one firm can influence the market price with its output decision. Recall that we have price-taking behavior under perfect competition which means that neither an individual consumer nor an individual producer can influence the market price. 

## Monopoly
A monopoly is an an industry that has only one producer of the product and there no close substitutes. We will see that a monopolist tends to reduce output in order to raise price. However, to increase the quantity sold, a monopolist must lower the price on each unit sold. The monopolist makes money on the additional quantity sold and loses money on the goods that it was already selling, as these are now sold at a lower price. Reasons for a monopoly to exit are usually barriers to entry.

- Economies of scale: One firm can operate at lower average cost than multiple firms and there are decreasing average cost over the entire market range. A natural monopoly can produce for the entire market and/or there may be large sunk cost and/or fixed cost.
- Legal barriers: Protection of intellectual property (e.g., copyrights, patents), control over a scarce resource, or barriers put in place by firms to discourage entry.
- Network externalities: Added benefits for all users of a good or service that arise because other people are using it too. Joining a large network is more beneficial than joining a small network. Think about Facebook and LinkedIn.

As opposed to perfect competition, there is the possibility of profits in the long-run. Examples of monopolies are Microsoft (to a certain extent), the IUPUI parking system, or De Beers. Groups can also act as a monopolist (cartel) such as OPEC, unions, or United Potato Growers of America. Note that a monopoly can create competition for itself in the case of durable goods monopolies. A durable good does not depreciate once it is on the market. Imagine that you are the owner of the only gold mine on Earth. At the beginning, you have a monopoly. But assuming that gold is never transformed into something else (e.g., like gasoline is burned), there will soon be a second hand gold market which competes with your monopoly. Another example of a durable monopoly that you can all relate to are textbooks.


### Monopoly Output: Overview
The monopolist will set marginal revenue equal to marginal cost, i.e., $MR=MC$, in order to find the profit maximizing output quantity. Recall that under perfect competition, we have marginal cost pricing, i.e., $p=MC$. It is important to understand that firm in a perfectly competitive market also sets $MR=MC$ but because the price is determined by the market, the marginal revenue from selling an additional unit is $p$ (Figure \ref{fig:MON_ProfitMax}). The following two example illustrate the monopolist's output decision under (1) constant and (2) increasing marginal cost. Note that a monopolistic market results in a deadweight loss. This is not caused by higher prices for consumers of higher profits for the monopolist but by the underproduction of goods.

```{r IPCprofitmax,echo=FALSE,fig.cap="Under perfect competition as well as under a monopoly situation, the profit maximizing output quantity is determined where marginal revenue is equal to marginal cost."}
revenue_mon    = function(x){100*x-2*x^2}
revenue_com    = function(x){p*x}
MR             = function(x){100-4*x} 
cost           = function(x){100+0.6*x^2}
mc             = function(x){1.2*x}
p              = 26
profit_mon     = optimr(3,function(x){(MR(x)-mc(x))^2},method=c("BFGS"))
profit_mon     = profit_mon$par
profit_com     = optimr(3,function(x){(p-mc(x))^2},method=c("BFGS"))
profit_com     = profit_com$par
par(mfrow=c(1,2))
  curve(revenue_com,from=0,to=50,xlab="output",main="Perfect Competition",yaxs="i",
        xaxs="i",ylim=c(0,1300))
       curve(cost,from=0,to=50,add=TRUE,yaxs="i",xaxs="i",col="blue")
       segments(profit_com,0,profit_com,revenue_com(profit_com),lty=2,col="black")
       segments(0,revenue_com(profit_com),profit_com,revenue_com(profit_com),lty=2,
                col="black")
       legend("bottomright",legend=c("Revenue","Total Cost"),fill=c("black","blue"),
              bty="n")
  curve(revenue_mon,from=0,to=50,xlab="output",main="Monopoly",yaxs="i",xaxs="i",
        ylim=c(0,1300))
       curve(cost,from=0,to=50,add=TRUE,yaxs="i",xaxs="i",col="blue")
       segments(profit_mon,0,profit_mon,revenue_mon(profit_mon),lty=2,col="black")
       segments(0,revenue_mon(profit_mon),profit_mon,revenue_mon(profit_mon),lty=2,
                col="black")
       legend("bottomright",legend=c("Revenue","Total Cost"),fill=c("black","blue"),bty="n")
rm(revenue_mon,revenue_com,MR,cost,p,profit_com,profit_mon,mc)
```

### Monopoly Output: Constant Marginal Cost
Assume a monopolist is faced with a demand that can be written as $Q=50-P/2$. The inverse demand function can thus be written as $P=100-2 \cdot Q$. The revenue can be written as
$$R = P \cdot Q = (100-2 \cdot Q ) \cdot Q = 100 \cdot Q - 2 \cdot Q^2$$
Marginal revenue is written as $MR = 100 - 4 \cdot Q$. Assume $MC =20$, then under perfect competition, we have $MC = P$, i.e., $20 = 100-2 \cdot Q$. And hence, the socially optimum output quantity is $Q_c=20$. Under monopoly pricing, we have $MC = MR$, i.e., $20 = 100-4 \cdot Q$. And hence, the monopoly output quantity is $Q_m=20$. This situation is illustrated in Panel (a) of the figure below.

```{r IPCmc,echo=FALSE,fig.cap="Examples of output decisions by a monopolist under constant and increasing marginal cost."}
demand         = function(x){100-2*x}
MR             = function(x){100-4*x}   
MC_con         = function(x){20+0*x}
MC_inc         = function(x){2*x}
q_con          = optimr(3,function(x){(MR(x)-MC_con(x))^2},method=c("BFGS"))
q_con          = q_con$par
q_inc          = optimr(3,function(x){(MR(x)-MC_inc(x))^2},method=c("BFGS"))
q_inc          = q_inc$par
par(mfrow=c(1,2))
  curve(demand,from=0,to=50,xlab="output",main="(a) Constant MC",yaxs="i",xaxs="i",col="black")
       curve(MR,from=0,to=50,add=TRUE,yaxs="i",xaxs="i",col="blue")
       curve(MC_con,from=0,to=50,add=TRUE,yaxs="i",xaxs="i",col="red")
       segments(q_con,0,q_con,demand(q_con),lty=2,col="black")
       segments(0,demand(q_con),q_con,demand(q_con),lty=2,col="black")
       legend("topright",legend=c("Demand","MR","MC"),fill=c("black","blue","red"),bty="n")
  curve(demand,from=0,to=50,xlab="output",main="(b) Increasing MC",yaxs="i",xaxs="i",col="black")
       curve(MR,from=0,to=50,add=TRUE,yaxs="i",xaxs="i",col="blue")
       curve(MC_inc,from=0,to=50,add=TRUE,yaxs="i",xaxs="i",col="red")
       segments(q_inc,0,q_inc,demand(q_inc),lty=2,col="black")
       segments(0,demand(q_inc),q_inc,demand(q_inc),lty=2,col="black")
       legend("top",legend=c("Demand","MR","MC"),fill=c("black","blue","red"),bty="n")
rm(q_con,q_inc,demand,MC_con,MC_inc)
```

### Monopoly Output: Increasing Marginal Cost}
Suppose that the cost function is written as $C = 100 + Q^2$ with increasing marginal cost function $MC = 2 \cdot Q$. If the inverse demand function is written as $P=100-2 \cdot Q$, then the marginal revenue function is $MR = 100 - 4 \cdot Q$. Similar to the example with constant marginal cost, we set $MR=MC$, i.e.,
$$100 - 4 \cdot Q = 2 \cdot Q$$
Solving leads to $Q_m=100/6$ which is illustrated in Panel (b), Figure \ref{fig:MON_MC}. What is the socially optimum output? Assume the government wants to achieve the socially optimal outcome with regulation. Further, assume that the government envisions taxing the monopolist. Does this solve the problem?

\subsection{Monopoly and Elasticity of Demand}
A monopolist produces on the elastic portion of the demand curve. It can be shown that
$$\underbrace{\frac{p-MC}{p}}_{\text{Lerner Index}} = -\frac{1}{\epsilon_D}$$
The interpretation of the above equation is that the more elastic demand the higher the availability of substitutes. Markup over marginal cost is reduced if the elasticity becomes larger (Figure \ref{fig:MON_elasticity}). Suppose a profit-maximizing monopolist is producing 800 units of output and is charging a price of \$40 per unit. If the price elasticity of demand for the product is -2, find the marginal cost of the last unit produced. Answer: \$20
$$\frac{p-MC}{p} = -\frac{1}{\epsilon_D} \quad \Leftrightarrow \quad p \left(1+\frac{1}{\epsilon_D} \right) = MC$$
```{r IPCelasticity,echo=FALSE,fig.cap="The magnitude of above marginal cost pricing depends on the elasticity of the demand function."}
demand_el      = function(x){100-x}
demand_in      = function(x){100-4*x}
MR_el          = function(x){100-2*x}   
MR_in          = function(x){100-8*x} 
MC             = function(x){2*x}
q_el           = optimr(3,function(x){(MR_el(x)-MC(x))^2},method=c("BFGS"))
q_el           = q_el$par
q_in           = optimr(3,function(x){(MR_in(x)-MC(x))^2},method=c("BFGS"))
q_in           = q_in$par
     par(mfrow=c(1,2))
          curve(demand_el,from=0,to=50,xlab="output",main="(a) Elastic Demand",yaxs="i",
                xaxs="i",col="black",ylim=c(0,100))
               curve(MR_el,from=0,to=50,add=TRUE,yaxs="i",xaxs="i",col="blue")
               curve(MC,from=0,to=50,add=TRUE,yaxs="i",xaxs="i",col="red")
               segments(q_el,0,q_el,demand_el(q_el),lty=2,col="black")
               segments(0,demand_el(q_el),q_el,demand_el(q_el),lty=2,col="black")
               legend("left",legend=c("Demand","MR","MC"),fill=c("black","blue","red"),bty="n")
          curve(demand_in,from=0,to=50,xlab="output",main="(b) Inelastic Demand",
                yaxs="i",xaxs="i",col="black",ylim=c(0,100))
               curve(MR_in,from=0,to=50,add=TRUE,yaxs="i",xaxs="i",col="blue")
               curve(MC,from=0,to=50,add=TRUE,yaxs="i",xaxs="i",col="red")
               segments(q_in,0,q_in,demand_in(q_in),lty=2,col="black")
               segments(0,demand_in(q_in),q_in,demand_in(q_in),lty=2,col="black")
               legend("top",legend=c("Demand","MR","MC"),fill=c("black","blue","red"),bty="n")
```

## Price Discrimination
A single-price monopoly is limited to charging the same price for each unit of output sold. If price discrimination is possible, the monopolist charges different prices to different customers, i.e., the monopolist can divide customers into different categories based on their willingness-to-pay (WTP). The firm must (1) have market power, (2) identify consumers' WTP, and (3) prevent low-price customers from reselling to high price customers in order for price discrimination to be feasible. Price discrimination is always beneficial to producers otherwise they would not charge different prices to different consumers. It can harm consumers because there will be a higher price for some consumers but it also has lower prices for others. The additional producer's profit represents a monetary loss to consumers. Examples are movie theaters, buses, trains, airplanes, amusement parks, or restaurants where a difference in prices is based on groups, e.g., students, seniors, children. Price discrimination can also be observed for events such as sport games where season tickets and quantity discounts are offered. There are three types of price discrimination:

- Perfect (or ``first degree'') price discrimination is achieved if the marginal revenue curve is equivalent to the demand curve.
- Nonlinear pricing or ``second degree'' price discrimination occurs with different types of consumers without observation of type. This results in offering different quantities of the good at different prices. Examples are different classes of service, hobbling a product (restrictions on the use of a device), quantity discounts, and so on.
- Imperfect (or ``third degree'') price discrimination occurs with different types of consumers with observation of type.

## Oligopoly
An oligopoly market is characterized by few sellers, blocked entry and exit, imperfect dissemination of information, and the opportunity for positive economic profits in the long-run. There is a strategic interdependence among the firms that is commonly represented by the Bertrand model, Cournot model or the Stackelberg model. A very powerful approach to model the strategic interaction among firms is game theory which will be covered in the next chapter.

The barriers to entry can result from (1) economies of scale or natural oligopoly, (2) reputation, (3) strategic barriers, or (4) legal barriers. In the case of economies of scale, one firm can supply a large part of the market (not all of it) which leads to lower per-unit cost than many small firms. Established oligopolists are likely to have favorable reputation as well. Strategic barriers can result from excess production capacity, market saturation, special contracts with distributors or from long-term arrangements with customers. As with the monopoly, legal barriers such as patents, copyrights, or zoning regulations can occur as well.

Oligopoly is a very common market structure. There can be a dominant single firm surrounded by a number of smaller firms (Anheuser-Busch 47\% and MillerCoors 30\%), three dominant firms (Universal Music Group 39\%, Sonic Music Entertainment 30\%, and Warner Music Group 19\%), duopoly which is an oligopoly with only two sellers (Airbus and Boeing). Examples of less concentrated markets are cereals with Kellogg, General Mills, Post, and Quaker Oats. But there are far more examples such as oil/gasoline, cars, cable/internet, smartphones, textbooks, rental cars, airlines, aluminum etc. at the national level or gasoline, food, and various professional services at the local level.

### Measurement of Market Concentration
The economic census provides a comprehensive statistical profile of the economy. Industry statistics are classified using the North American Industry Classification System (NAICS). The main categories are Agriculture, Forestry, Fishing and Hunting (Code 11), Mining, Quarrying, and Oil and Gas Extraction (Code 21), Utilities (Code 22), Construction (Code 23), Manufacturing (Code 31-33), Wholesale Trade (Code 42), Retail Trade (Code 44-45), Transportation and Warehousing (Code 48-49), Information (Code 51), Finance and Insurance (Code 52), Real Estate and Rental and Leasing (Code 53), Professional, Scientific, and Technical Services (Code 54), Management of Companies and Enterprises (Code 55), Administrative and Support and Waste Management and Remediation Services (Code 56), Educational Services (Code 61), Health Care and Social Assistance (Code 62), Arts, Entertainment, and Recreation (Code 71), Accommodation and Food Services  (Code 72), Other Services (except Public Administration)  (Code 81), Public Administration (Code 92).

NAICS  | Firms | Sales in billion |  CR$_4$ |  CR$_8$ |  CR$_{20}$ |  CR$_{50}$ | HHI 
-------|-------|------------------|---------|---------|------------|------------|--------
336992 | 61    | 10,988           |  81.8   |  93.7   |  98.0      |  99.9      | 2,477.1 
311111 | 199   |  14,505          |  71.0   |  83.5   |  92.6      |  98.2      | 2,325.1
323114 | 6,041 |  3,263           |  3.8    |  5.8    |  10.3      |  17.4      | 8.4
327320 | 2,460 |  34,837          |  22.6   | 28.2    |  39.9      |  52.9      | 312.9

        \caption{2007 Economic Census Example: Military armored vehicle, tank, and tank component manufacturing (336992), Dog and cat food manufacturing (311111), Quick printing (323114), Ready-mix concrete manufacturing (327320)}
    \end{table}

Concentration ratios measure the combined market share percentage of the i$^{\text{th}}$ leading firms:
$$CR_n = \sum_{i=1}^n \frac{{\text{Firm Sales}}_i}{\text{Industry Sales}} \times 100$$

For a monopoly, we have $CR_n = 100$ and for a perfectly competitive industry we have $CR_n \approx 0$. So if we have many firms, the concentration ratio decreases. The Herfindahl-Hirschmann Index is the sum of squared market share percentage for all competitors:
$$ HHI = \sum_i \left( \frac{{\text{Firm Sales}}_i}{\text{Industry Sales}} \times 100 \right)^2$$
For a monopoly, $HHI=10,000$ and for a perfectly competitive industry $HHI \approx 0$. %Suppose we have four firms in the market

$F_1$   | $F_2$ | $F_3$ | $F_4$ |  $CR_2$   | $CR_4$    |  HHI 
--------|-------|-------|-------|-----------|-----------|-----
25      | 25    | 25    | 25    |  50       | 100       | 2500 
70      | 10    | 10    | 10    |  80       | 100       | 5200 
40      | 40    | 10    | 10    |  80       | 100       | 3400 


### Bertrand, Cournot, and Stackelberg Models
The following models attempt to explain competitive behavior among two firms or more firms. All three models are relatively old. The Cournot model was published in 1838, the Bertrand model was formulated in 1883, and the Stackelberg model was described in 1934.
% Cournot example: Real estate in a large metropolitan area. say something about product differentiation. betrand model can be for airlines.
Suppose that the demand is written as $P=30-Q=30-Q_1-Q2$ and the marginal cost is equal to 2. Then the profit is written as
$$\pi = 30-Q_1-Q_1^2-Q1 \cdot Q2 - 2 \cdot Q_1$$
The reaction function is written
$$Q_i = 14-\frac{Q_j}{2}$$
and the price is $P=11.34$. In the Stackelberg model, the leading firm moves first and the follower observes the action of the leading firm and decides on output quantity.

### Collusion
Recall the Cournot output of $q = 9.34$ and that under perfect competition, we have $p=MC$ leading to $q_1 = q_2 = 14$ and $\pi_1 = \pi_2 = 0$. The question becomes if the two firms could do better for themselves. The answer is yes if both firms act together as a monopolist with $p = 30 - q$, $MR = 30 - 2q$, $q = 14 \Rightarrow q_1 = q_2 = 7$.

Firms can agree on explicit collusion where cooperation involves direct communication between competing firms about setting prices. This type of behavior is usually illegal and involves severe penalties. Tacit collusion is oligopolistic cooperation which does not involve an explicit agreement. To achieve this, there is a tit-for-tat strategy. Do to another player this period what he has done to you in the previous period. Limits to collusion are the downward sloping demand function, legal issues, and incentives to cheat on any agreement because each player can do even better by cheating.


Suppose a simple demand function $p = 1-q$ and no cost. The profit of firm 1 is $\pi = q(1-q) = q-q^2$. Under a monopoly, we have $q = 0.5$ and $\pi = 0.25$. If a second firm is in the market, we have
    \begin{align*}
        \pi_1 & = q_1(1-q_1-q_2)\\
        \pi_2 & = q_2(1-q_1-q_2)\\
              & = q_2 - q_1 q_2 - q^2
    \end{align*}
Optimal quantity for firm 2
    \begin{equation*}
        q_2 = \frac{1-q_1}{2}
    \end{equation*}
Reaction of firm 1
    \begin{align*}
        \pi_1  & = q_1(1-q_1-\frac{1-q_1}{2})\\
                & = q_1 - q_1^2 - \frac{q_1}{2} + \frac{q_1^2}{2}
    \end{align*}


% Go over the Binger and Hoffman example about price discrimination.

## Examples
The United Potato Growers of America ``aspires to be to potatoes what OPEC is to oil by carefully managing supply to keep demand high and constant, resulting in a more stable return for farmers.'' This is how the Wall Street Journal describes the situation in the U.S. potatoes market.\footnote{``This Spud's Not for You: Growing Co-Op of Farmers Seeks to Become OPEC of Potatoes by Controlling Supply, 26 September 2006.} Agricultural producers are exempt from anti-trust legislation (the Capper-Volstead Act exempted farmers from federal antitrust laws) which allowed the United Potato Growers of America to destroy 680 million pounds of potatoes in 2005 in order to keep prices high. This increased the price of potatoes by 48.5\% which was mostly absorbed by retailers and thus, consumers did not see a large price increase. There has been an issue of overproduction of potatoes which resulted in a decrease of price. Controlling the supply helped stabilize (or increase) the price of potatoes.

% check or search for OXFAM behind the brand infographic for the original source of the food oligopoly.
% https://www.oxfamamerica.org/explore/stories/these-10-companies-make-a-lot-of-the-food-we-buy-heres-how-we-made-them-better/
% https://worldmaritimenews.com/archives/272206/shipping-majors-cleared-in-us-cartel-investigation/

## Exercises

1. *Drug Monopoly*: Mark is a producer who has the monopoly on a drug curing a particular disease. The following equations characterize his economic environment:

    - Demand:
        $$Q=150-3 \cdot P \Leftrightarrow P=50-\frac{Q}{3}$$
                    \end{equation*}
    - Total revenue:
                    \begin{equation*}
                        TR = 50 \cdot Q - \frac{Q^2}{3}
                    \end{equation*}
    - Marginal revenue:
                    \begin{equation*}
                        MR = 50-\frac{2}{3} \cdot Q
                    \end{equation*}
    - Total cost:
                    \begin{equation*}
                        TC= 4 \cdot Q
                    \end{equation*}
    - Marginal cost
                    \begin{equation*}
                        MC=4
                    \end{equation*}
            \end{itemize}

What is the profit maximizing price and output quantity? What is his profit? What is the efficient price and output quantity? Calculate the deadweight loss associated with the monopoly situation? To regulate the monopoly, a tax of \$1 is imposed. Given the tax, calculate the new price, quantity, and deadweight loss.

2. *Electricity Market Monopoly*: An electric utility company has a monopoly in a large metropolitan area. The inverse demand and marginal revenue functions are $P=50-Q$ and $MR= 50-2 \cdot Q$. The marginal cost function is $MC = 10+2 \cdot Q$. What is the profit maximizing price and output quantity? What is the efficient price and output quantity? Calculate the deadweight loss associated with the monopoly situation? The government sets a price ceiling at $p=MC$. What is the price and output of the regulated monopoly?
        
3. *Monopoly Profit Tax*: Assume a regulator wants to impose a 25\% tax on the profit of a monopolist to reduce the inefficiency associated with the monopolist. What is the effect of this policy on the deadweight loss. Use a graph to illustrate.