# Introduction
```{r INTfrontmatter,echo=FALSE,message=FALSE,warning=FALSE}
load("pme.RData")
library(plot3D)
library(optimr)                                 
library(ggplot2)
library(gridExtra)      
library(pracma)
library(kableExtra)
```

Economics analyzes how individuals, households, firms, governments, and society in general employ scarce resources to produce goods and services. Economists are concerned about the efficient allocation of resources. That is, could resources be reallocated in such a way to make society better off. Economists use mathematical, theory-based models to represent the aforementioned actors and how they interact in a market or some other economic activity. Only relevant features are included in models and not every detail. Empirical work is done to test economic theories. Mathematics is a necessary tool for economists. Although this book limits the use of mathematics, the reader should keep in mind that behind that there is a mathematical relationship behind every graph presented. 

Economic concepts are used (explicitly or implicitly) by individuals, firms, and governments. An economist would say that most of your everyday life decisions are based on economic principles starting with the decision to get out of bed in the morning. The American Economic Association (AEA) has compiled a [list of academic papers](https://www.aeaweb.org/resources/teachers/economics-concepts-examples-in-research) that help explain economic concepts using real world examples.

Microeconomics and macroeconomics are the two large branches in economics. Microeconomics analyzes the behavior of individuals and firms whereas macroeconomics assess the behavior of the economy as a whole. Important topics in macroeconomics are unemployment, interest rates, trade, national income, inflation, and many more. This book covers microeconomics which includes the following topics:

- **Consumer theory**: Consumer theory uses the concept of utility maximization to derive the demand curve. Consumer theory is important in terms of welfare economics, e.g., what are the welfare effects for the consumer after a policy change.
- **Producer theory**: Whereas the goal of consumer theory is to derive the demand function for a good, producer theory derives the supply function for a good. Production theory is less abstract than consumer theory. Terms such as profit maximization and cost minimization are easier to understand. 
- **Demand and supply**: The combination of demand and supply determines the quantity and price which clear the market and lead to a market equilibrium, i.e., a situation with neither excess demand nor excess supply. It is important to differentiate between a change in quantity demanded (supplied) versus a change in demand (supply). There are many applications of demand and supply such as analyzing the effects of the 2012 drought in the Midwest on crop prices or how an increase in carbon fiber demand for wind farms has spill-over effects on the market for high-end bikes. The concept of elasticity is introduced in the context of demand and supply as well.
- **Perfect competition**: Once the supply and demand model is developed, we will focus on different market structures. The first such structure is perfect competition that is characterized by price-taking behavior of the market actors, i.e., consumers and producers.
- **Imperfect competition**: As opposed to a perfectly competitive market where producers are price takers, imperfect competition is characterized by price-setting behavior. If there is one producers for a particular good, we talk about a monopoly. A small number of producers is called oligopoly. Examples of oligopoly are widely available such as Airbus and Boeing or car manufacturing.
- **Game theory**: The strategic interaction among economic agents is analyzed with game theory. Each game has three components: (1) players, (2) actions, and (3) payoffs. Examples of games are auctions, bargaining, football, or chess.
- **Public Policy**: Besides imperfect competition, there are many other examples where the market fails to provide the efficient outcome. For example, most economic analysis requires the existence of a market. There are many goods for which no market exists. Many of them are found in environmental economics such as clean water and clean air. We are going to look at the implications of the absence of markets and the limitations to what people call the free market, e.g., [Gulf of Mexico Hypoxic Zone](https://oceanservice.noaa.gov/facts/deadzone.html). Another example of market failure is the presence of asymmetric information. The used car market is a good example because two parties enter a contract where one party (buyer) is not fully informed. For example, we will also learn why dressing nicely for a job interview is important or why you should post as many photos as possible when you sell your car.
- **Risk and uncertainty**: Almost all activities that are conducted by consumers or firms involve some sort of uncertainty. Decisions that are made today, affect the outcome tomorrow that is usually uncertain. How do individuals and firms make decision given those uncertainties.

There are many economic concepts that are useful in everyday life. Those concepts are analyzed in detail in the subsequent chapters. Some of the most important concepts are the following:

- **Marginal Analysis**: Economists focus on the margin, i.e., what is the benefit of the next good that is bought or sold, which leads to the concept of marginal analysis. Once you have purchased something, what matters is what you will do next. Can you make yourself any better? Our goal is to maximize total net benefit: the value of the good minus the cost. For a proposed action: Additional (marginal) benefit must be bigger or equal the additional (marginal) cost. Examples are choosing the production quantity or renting a car. For the maximization of total benefit, we focus on marginal benefit and marginal cost. If the marginal benefit is larger than the marginal cost, total benefit will increase. If the marginal cost is larger than the marginal benefit, total benefit will decrease. The only time that total benefit will not rise or fall is when marginal benefit is equal to marginal cost. This is where total net benefit is maximized.
- **Sunk Cost**: A concept closely related to marginal analysis are sunk costs. Sunk costs are expenditures that have been made and cannot be recovered. Following from marginal analysis, sunk costs should be ignored. Since sunk costs cannot be changed, they should not influence decision-making. This concept is controversial because it does not always hold in reality.
- **Opportunity Cost**: Another example of costs that are considered by economists are opportunity costs which are the value of the next best alternative use of a resource, i.e., cost of forgone opportunities. For MPA students, the cost of going to the O'Neill School is tuition plus the opportunity cost, i.e., salary you forgo by not working. Opportunity costs are important to consider, but often ignored. Opportunity costs relate to the key concept of scarcity. Once a resource has been used, it cannot be used for something else.

Economics should be thought off in very broad terms and not just money and/or profit. Consider the example of saving human lives in the context of transport planning. In many regions in the U.S. and the world, people have a choice when it comes to transport modes (e.g., car, bike, train). Depending on the price of those modes, consumers substitute one for the other. There are also different fatality rates associated with each transport mode with cars probably having one of the highest. So if you increase train ticket prices to finance safety investments, people shift from the train to the road potentially causing more fatalities despite the investment in safety. Just focusing on road safety, an increase in the price of gasoline reduces road fatalities ([Burke and Nishitateno, 2014](https://doi.org/10.1111/ecin.12171)).

## Real versus Nominal Prices
Before starting with consumer theory, let us first look at indices and the difference between nominal and real prices. An index number assigns a single value to several individual numbers in order to quantify trends. An index is series of numbers used to track a variable's rise and fall over time. Examples are the Consumer Price Index (CPI), the Dow Jones Industrial Average, the S\&P 500, or your GPA. The CPI is important because it allows us to differentiate between real and nominal prices. In general, only real prices are important. Let us illustrate this point with an example.

Suppose that your income in time period 1 is \$100. The price of apples and milk is \$5 and \$10, respectively. You purchase 4 apples and 8 gallons of milk. In time period 2, your income is \$120 and the price of apples and milk increased to \$6 and \$12, respectively. Has your nominal income changed? Has your real income changed? What about an income of \$110? Many people compare 1970 gasoline prices to today and complain about the price increase without realizing that they are comparing nominal prices. This would be similar to complaining about the cost increase of a Rolls-Royce. In 1921, a Rolls-Royce Silver Ghost cost \$12,000 which at the time was equivalent to the price of three Cadillac.

Besides the CPI, there are two other indices that can be used to differentiate real from nominal prices: GDP deflator and Producer Price Index (PPI). The GDP deflator includes goods and services purchased by businesses, governments, and foreigners whereas the CPI includes the price changes for consumer goods and services. The PPI includes goods and services purchased by producers and firms. For example, the price a large industrial firm pays for electricity is different, and behaves differently, than the price consumers pay.

```{r INTcpi,echo=FALSE,fig.cap="CPI for [All Items in U.S. City Average](https://fred.stlouisfed.org/series/CPIAUCSL), [Medical Care in U.S. City Average](https://fred.stlouisfed.org/series/CPIMEDSL), and [Tuition, Other School Fees, and Childcare in U.S. City Average](https://fred.stlouisfed.org/series/CUSR0000SEEB)"}
ggplot(data=cpi,aes(x=Year,y=value,group=Legend))+geom_line(aes(linetype=Legend),size=0.5)+
        theme_bw()+theme(legend.position="bottom")+ylim(0,800)+
        ylab("CPI (1982-1984=100)")+theme(axis.title.x=element_blank())
rm(cpi)
```

The figure above illustrates the evolution of the CPI since 1978. The CPI is a "measure of the average change over time in the prices paid by urban consumers for a market basket of consumer goods and services" (Bureau of Labor Statistics). The CPI is a basket of about 80.000 items. The weights in the basket are food and beverages (14.99\%), housing (42.69\%), apparel (3.76\%), transportation (17.25\%), medical care (6.27\%), recreation (5.54\%), education and communication (6.03\%), and other goods and services (3.47\%). Urban consumers represent about 87\% of the U.S. Population. The CPI tracks and reports price level. In this class, we measure all prices in real terms. We can use the following equation to calculate the CPI:
$$CPI_t = \frac{MB_t}{MB_b} \cdot 100$$
where $MB_t$ and $MB_b$ represents the cost of the market basket in time period $t$ and the base year $b$, respectively. The CPI base year is 1982 to 1984 but because it is an index, the base year can be adjusted for specific purposes. The CPI is also used as an economic indicator to set policy targets or to adjust dollar values such as pensions. It is also used as a deflator for price series to translate nominal into real prices. For example, the figure below shows the nominal and the real price of gasoline. We can see that in 2000, the nominal price of gasoline was around \$1.50 but the real price was over \$2.00. For real and nominal energy prices going back to 1970, see the [Real Prices Viewer](https://www.eia.gov/outlooks/steo/realprices/) from the U.S. Energy Information Administration. 

```{r INTgasoline,echo=FALSE,fig.cap="Real and nominal [US Regular All Formulations Gas Price](https://fred.stlouisfed.org/series/GASREGM). The [GDP Implicit Price Deflator in United States](https://fred.stlouisfed.org/series/USAGDPDEFAISMEI) is used to deflate the prices. Source: Federal Reserve Economic Data."}
ggplot(data=gasoline,aes(x=Year,y=value,group=Legend))+geom_line(aes(linetype=Legend),size=0.5)+
        theme_bw()+theme(legend.position="bottom",axis.title.x=element_blank())+ylim(0,5)+
        ylab("Price in $ per Gallon")
rm(gasoline)
```

The constant dollar calculation can be written as:
$$CD_t = \frac{ND_t \cdot CPI_b}{CPI_t}$$
where $CD_t$ represents the constant dollar or real value, $ND_t$ is the nominal value in period $t$, and $CPI_b$ and $CPI_t$ are the Consumer Price Index in the base period and period $t$, respectively. For example, assume the following parameters: $ND_{1979} = 0.84$ (May 1979), $CPI_{1979} = 71.4$ ($CPI = 100$ in 1982-1984), $CPI_{2012} = 228.8$, $b = 2012$ and $t = 1979$. Plugging those values into the equation for the constant dollar calculation and we get:
$$CD_{2012} = \frac{ND_{1979} \cdot CPI_{2012}}{CPI_{1979}} = \frac{0.84 \cdot 228.8}{71.4} = 2.69$$
Let us consider a second example using the \$12,000 Rolls-Royce Silver Ghost. Let us use January 1921 and June 2020 as the two time points of interest.
$$CD_{2020} = \frac{ND_{1921} \cdot CPI_{2020}}{CPI_{1921}} = \frac{12,000 \cdot 257.797}{19.0} = \$162,819.20$$

Figure \ref{fig:INT_infueltax} also shows that an increasing price series in nominal terms can actually be decreasing in real terms. The fuel tax revenue in other states and at the federal level evolve similar to the one in Indiana. So in case you were wondering where the potholes come from.


```{r INTinfueltax,echo=FALSE,fig.cap="Real and nominal fuel tax revenue for Indiana. The [GDP Implicit Price Deflator in United States](https://fred.stlouisfed.org/series/USAGDPDEFAISMEI) is used to deflate the prices. Source: [State Government Tax Collections, Motor Fuels - Selective Sales Taxes in Indiana](https://fred.stlouisfed.org/series/INFUELTAX) from the Federal Reserve Economic Data."}
ggplot(data=infueltax,aes(x=Year,y=value/1000,group=Legend))+
        geom_line(aes(linetype=Legend),size=0.5)+theme_bw()+
        theme(legend.position="bottom",axis.title.x=element_blank())+ylab("Revenue in Million $")+
        ylim(0,1500)
rm(infueltax)
```

There are some problems with the CPI. The substitution bias refers to the situation where consumers shift their purchases away from goods whose relative prices rise towards lower priced goods. Additional problems are associated with new, higher priced goods that replace older goods, failing to take into account of quality improvements that raise prices, and the growth in discounting, i.e., people switching to low-cost discount stores.

The inflation rate is the percentage change in the price level from one period to the next. The inflation rate can be calculated as follows:
$$\pi_t=\frac{CPI_t-CPI_{t-1}}{CPI_{t-1}}\cdot 100$$
```{r INTinflation,echo=FALSE,fig.cap="Inflation as the growth rate from the previous period of the [Consumer Price Index: Total All Items for the United States](https://fred.stlouisfed.org/series/CPALTT01USA657N). Source: Federal Reserve Economic Data."}
ggplot(data=inflation,aes(x=Year,y=value))+geom_line(size=0.5)+
        theme_bw()+ylim(-5,15)+ylab("Inflation Rate")+theme(axis.title.x=element_blank())
rm(inflation)
```
## Exercises

1. *Real Gasoline Price*: For this exercise, you will calculate the price of gasoline that prevailed several month ago in terms of the most recent month the CPI is available. First, go to [Short-Term Energy Outlook: Real Price Viewer](https://www.eia.gov/outlooks/steo/realprices/), select "Regular Gasoline Retail Prices" as the Price Series and "Monthly" as the Period Type, and download the data. Second, download the monthly CPI data for "All Urban Consumers" from the Bureau of Labor Statistics (BLS) webpage. Next, take the last two digits of your student ID number. That number determines how many months you go back to find the gasoline price I want you to use for this exercise. For example, if the last two digits are 08, then you go back 8 months. Calculate the gasoline price that prevailed 8 months ago in terms of dollars of the most recent month. In your answers, (1) report the month and year applicable to you, (2) the CPI in that month, (3) the gasoline price in that month, (4) the most recent CPI value, and (5) the real gasoline price.
2. *Annual State Government Tax Collections*: There are extensive statistics on state finances, i.e., revenue and expenditures. The Federal Reserve Bank of St. Louis maintains the Federal Reserve Economic Data (FRED) database which contains state finance statistics among thousands of other data series. For this exercise, go to [Annual State Government Tax Collections](https://fred.stlouisfed.org/release/tables?rid=143&eid=151912) and pick a state of your choice. Next, pick one of the following revenue categories: (1) General Sales and Gross Receipts Taxes, (2)  Motor Fuels Sales Tax, (3) Motor Vehicle License, (4) Individual Income Taxes, or (5) Corporation Net Income Taxes. Download the data (i.e., state and revenue source over multiple years) to an Excel file. The data is in nominal terms and your tasks for this exercise are the following:
    a. Download the annual data for the [GDP deflator](https://fred.stlouisfed.org/series/USAGDPDEFQISMEI). Make sure to use the orange button "EDIT GRAPH" to get the annual data.
     b. Calculate the real tax revenue for the time series and state chosen. The base year is the most current year. In your answer, include a graph which shows the evolution of the tax revenue in real and nominal terms.
     c. Did the tax revenue increase or decrease in nominal terms?  Did the tax revenue increase or decrease in real terms? Given the size of the revenue source as a share of overall tax revenue, what are the implications?