# Producer Theory
The goal of producer theory is to derive the supply curve. We will first look at production functions and profit maximization. In the subsequent chapter, we will talk about cost minimization which ultimately leads to the supply function. 

There is a YouTube video associated with this chapter:

- [Producer Theory](https://youtu.be/EilB16BfgG0)

## Production Functions
Production is the process of combining inputs to make goods and services. A production function indicates the maximum amount of output a firm can produce over some period of time from each combination of inputs. Inputs can be hours of labor, \$ of capital, cubic feet of natural gas, or tons of raw material. There are similarities between consumer theory and producer theory:

- Preferences $\Leftrightarrow$ Production technology
- Budget constraint $\Leftrightarrow$ Cost constraints
- Consumption choices $\Leftrightarrow$ Production choices
- Indifference curves and budget lines become isoquants and isocost lines

Production functions can be thought of in a very broad context, e.g., studying for an exam or running a 100 meter world record (9.58 seconds). Consider Panel (a) "Fertilizer Application" in the figure below. In the graph, we have yield as a function of nitrogen application: $y = f(N)$. This application could be of interest if you consider the cost associated with water quality and water treatment plants, e.g., Eagle Creek.

```{r PROferteff,echo=FALSE,fig.cap="Production functions: Panel (a) on the left shows crop yield as a function of fertilizer application. Panel (b) on the right illustrates the concept of inefficient and unattainable production points."}
yield = function(N){2.24*N-0.0056*N^2}
par(mfrow=c(1,2))
  curve(yield,from=0,to=200,ylab="Yield",xlab="Application rate in lbs. per acre",
        main="Fertilizer Application",yaxs="i",xaxs="i")  
  curve(yield,from=0,to=200,ylab="Output",xlab="Input",
        main="Efficient and Inefficent Points",yaxs="i",xaxs="i")
  points(x=c(100,100),y=c(100,200))
  segments(100,0,100,200,lty=2)
  segments(0,100,100,100,lty=2)
rm(yield)
```

## Production in the Short- and Long-Run
In producer theory, we differentiate between the short-run and the long-run. In the short-run, some inputs cannot be adjusted and they are so-called fixed inputs. Examples of fixed inputs are machinery and factories. We classify fixed input as capital ($K$). The short-run production function is written as $q = f(\bar{K},L)$. The long-run is a time horizon long enough for a firm to vary all of its inputs. Variable inputs can be adjusted up or down as the quantity of output changes. Consider production in the short run with labor ($L$) as the only input. The total product is the maximum quantity of output that can be produced from a given combination of inputs. The marginal product of labor is the additional output produced if one more worker is hired, i.e.,
$$MPL= \frac{\Delta Q}{\Delta L}$$
And the average product is defined as the average quantity per worker
$$AP= \frac{Q}{ L}$$
The law of diminishing marginal product says that as more and more of any input is added to a fixed amount of other inputs, its marginal product will eventually decline. Note that for low inputs, we can have an increasing marginal product to labor, i.e., MPL increases as more labor is hired. But in most cases, we are faced with diminishing marginal product of labor. Examples that you find in everyday life are studying for an exam or training for a marathon.

Returns to scale can be either constant, increasing, or decreasing.

- Constant returns to scale: $t \cdot f(x) = f(t \cdot x)$
- Increasing returns to scale: $t \cdot f(x) < f(t \cdot x)$
- Decreasing returns to scale: $t \cdot f(x) > f(t \cdot x)$

The figure below illustrates the relationship between production function, average product, and marginal product. The production function depicted also exhibits increasing marginal product due to specialization at the beginning and then decreasing marginal returns.

```{r PROapmp,echo=FALSE,fig.cap="Relationship between production function, average product, and marginal product",fig.height=12,fig.width=6}
profun    = function(x){x+4*x^2-0.03*x^3}
average   = function(x){(x+4*x^2-0.03*x^3)/x}
marginal  = function(x){1+8*x-0.09*x^2}
par(mfrow=c(2,1))
  curve(profun,from=0,to=90,ylab="Output",xlab="Input",main="Production Function",yaxs="i",xaxs="i")  
  curve(marginal,from=0,to=90,ylim=c(0,200),ylab="MP and AP",xlab="Input",
        main="Average Product and Marginal Product",col="red",yaxs="i",xaxs="i")
  curve(average,from=0,to=90,add=TRUE,yaxs="i",xaxs="i")
  legend("bottom",legend=c("Average Product","Marginal Product"),fill=c("black","red"))
rm(average,marginal,profun)
```

The more interesting case is production in the long-run where all inputs are variable. So in this section, we consider the inputs labor ($L$) and capital ($K$). The figure below shows a so-called Cobb-Douglas production function for the case of two inputs.

```{r PROcobbdouglas,echo=FALSE,fig.cap="Production function with two inputs."}
x    = as.numeric(seq(1,50))
y    = as.numeric(seq(1,50))
grid = mesh(x,y)
z    = with(grid,x^0.2*y^0.4)
par(mfrow=c(1,2))
    persp3D(z=z,x=x,y=y,contour=list(side=c("zmax","z")),zlim=c(0,12),phi=30,theta=20,d=10,box=FALSE)
    image2D(z=z,x=x,y=y,contour=TRUE)
rm(x,y,z,grid)
```

The Marginal Rate of Technical Substitution (MRTS) is the rate at which a firm can substitute one input for another while keeping output constant. The MRTS decreases as we move rightward along an isoquant. It is also the slope of the isoquant.
$$MRTS=-\frac{\Delta K}{\Delta L}=\frac{MP_L}{MP_K}$$
Recall that
$$MP_K = \frac{\Delta Q}{\Delta K} \quad \text{and} \quad MP_L = \frac{\Delta Q}{\Delta L}$$
Every firm wants to minimize its costs:
$$C = p_L \cdot L + p_K \cdot K \\
  TC = w \cdot L + w \cdot K \\
  K = \frac{TC}{r} - \frac{w}{r} \cdot L$$

The optimal choice is the point where an isocost line is tangent to the isoquant for that output level
$$MRTS= -\frac{\Delta K}{\Delta L} = \frac{p_L}{p_K}\\
        \frac{MP_L}{p_L} = \frac{MP_K}{P_K}$$
For any number of inputs: The marginal product per dollar of any input will be equal to the marginal product per dollar of any other input.

```{r PROexpansionpath,echo=FALSE,fig.cap="Short-run and long-run expansion path"}
xaxis          = c(0,8)
yaxis          = c(0,8)
a              = 0.5
b              = 0.5
w              = 2
r              = 2.5
Lstar          = function(Q){return((Q^(1/(a+b)))*((b*r)/(a*w))^(a/(a+b)))}
Kstar          = function(Q){return((Q^(1/(a+b)))*((a*w)/(b*r))^(b/(a+b)))}
isocost        = function(L){return(TC/r-(w/r)*L)}
isoquant       = function(L){return((Q/L^b)^(1/a))}
exppath        = function(L){return((a*w*L)/(b*r))}
     Q         = 2
     Loptimal  = Lstar(Q)
     Koptimal  = Kstar(Q)
     TC        = Loptimal*w+Koptimal*r
     curve(isoquant,from=xaxis[1],to=xaxis[2],yaxs="i",xaxs="i",ylim=yaxis,ylab="K",xlab="L")
     curve(isocost,from=xaxis[1],to=xaxis[2],yaxs="i",xaxs="i",add=TRUE)
     Q         = 3
     Loptimal  = Lstar(Q)
     Koptimal  = Kstar(Q)
     TC        = Loptimal*w+Koptimal*r
     curve(isoquant,from=xaxis[1],to=xaxis[2],yaxs="i",xaxs="i",ylim=yaxis,add=TRUE)
     curve(isocost,from=xaxis[1],to=xaxis[2],yaxs="i",xaxs="i",add=TRUE)
     Q         = 4
     Loptimal  = Lstar(Q)
     Koptimal  = Kstar(Q)
     TC        = Loptimal*w+Koptimal*r
     curve(isoquant,from=xaxis[1],to=xaxis[2],yaxs="i",xaxs="i",ylim=yaxis,add=TRUE)
     curve(isocost,from=xaxis[1],to=xaxis[2],yaxs="i",xaxs="i",add=TRUE)
     curve(exppath,from=xaxis[1],to=xaxis[2],yaxs="i",xaxs="i",add=TRUE)
rm(a,b,exppath,isocost,isoquant,Koptimal,Kstar,Loptimal,Lstar,Q,r,TC,w,xaxis,yaxis)
```

## Profit Maximization
Assume we have a single output that uses workers. The output price is $p$ and the wage per worker is $w$. Then the profit maximization problem can be written as:
$$\max \quad p \cdot f(L) - w \cdot L$$
And the solution to this problem can be written as:
$$p \cdot f'(L) = w \quad \Leftrightarrow \quad f'(L) = \frac{w}{p}$$
Put differently $\pi = p \cdot y - w \cdot x$. Solving for $y$:
$$y =\frac{\pi - w \cdot x}{p} = \frac{\pi}{p} + \frac{w}{p} \cdot x$$
The profit maximization problem as a function of output is $\pi(q) = R(q) - C(q)$ where $R = p \cdot q$. To find the profit maximizing output, we need the marginal revenue
$$\frac{\Delta R}{\Delta q} = p$$
and the marginal cost
$$\frac{\Delta C}{\Delta Q} = MC$$
Hence, the profit maximization condition is: $MC(q) = MR(q)$. This is true for any market structure. What differs across market structures is marginal revenue.

```{r PROprofitmax,echo=FALSE,fig.cap="Two different ways of illustrating profit maximization"}
p              = 164
cost           = function(q){return(450+100*q-3*q^2+0.2*q^3)}
mc             = function(q){return(100-6*q+0.6*q^2)}
profitzero     = function(q){return((p-mc(q))^2)}
revenue        = function(q){p*q}
profit         = function(q){revenue(q)-cost(q)}
qstar          = optim(par=p,profitzero,method="Brent",lower=0,upper=1e6)
qstar          = qstar$par
profit         = revenue(qstar)-cost(qstar)
tangent        = function(q){-profit+mc(qstar)*q}     
par(mfrow=c(1,2))
  curve(revenue,from=0,to=25)
  curve(cost,add=TRUE)
  curve(tangent,from=0.75*qstar,to=1.25*qstar,add=TRUE)
  a              = 2
  b              = 0.5
  w              = 3
  p              = 6
  profun         = function(L){a*L^b}
  Lstar          = (w/(p*a*b))^(1/(b-1))
  profit         = p*profun(Lstar)-Lstar*w
  tangent        = function(L){profit/p+w/p*L}
  curve(profun,from=0,to=10)
  curve(tangent,add=TRUE)
rm(p,cost,mc,profitzero,revenue,profit,qstar,tangent,a,b,w,profun,Lstar)
```

## Exercises
1. ***Nitrogen Use*** (*): A farmer uses nitrogen ($N$) as an input to produce corn yield as the output. The production function which relates nitrogen inputs to corn yield is written as $y=f(N)$. With the current management practices, the farmer obtains $y^\star$ bushels of corn with $N^\star = 120$ pounds of nitrogen fertilizer. Given that setup, we have $y^\star < f(120)$. Are the inputs used efficiently? Use a graph to justify your answer.

2. ***Malthus*** (*): The [Malthusian Catastrophe](https://en.wikipedia.org/wiki/Malthusianism) was the prediction by the English cleric and scholar Thomas Malthus (1766-1834) that population will grow faster than agricultural production. Use the concept of diminishing returns to labor in the short-run to explain his reasoning. Explain why it did not happen. Use a graph to justify your answer.

4. ***Profit Function*** (**): The production function of a firm is written as $f(x) = 8 \sqrt{x}$ where $x$ are the units of input. The per-unit output price is \$100 and the per-unit input cost is \$75. Write down the profit only as a function of $x$.

5. ***Electricity Producer*** (***): Assume an electricity producer that currently uses a mix of natural gas and coal as inputs to produce electricity as depicted in the figure below. In the short-run, the electricity producer can adjust neither coal nor natural gas as the input for electricity production, i.e., both inputs are fixed. The initial input mix of coal and natural gas is depicted in the figure belwoas well as the current isoquant. The output does not change throughout the exercise. 
    - Reproduce the figure and sketch an isocost for the quantity of electricity produced.
    - Assume that in the short-run, the price of coal decreases relative to natural gas. What happens to the isocost line? Is the choice of inputs optimal after the price change? Justify your answer.
    - In the long-run, what will happen to the combination of coal and natural gas used in the production of electricity if the price of coal remains low. Illustrate in your graph.
```{r PROelectricity,echo=FALSE,fig.cap="Electricity production with natural gas and coal."}
a                   = 0.3
b                   = 0.5
A                   = 100
Q                   = 300
isoq                = function(x){(Q/(A*x^a))^(1/b)}    
x1                  = 1.5
x2                  = isoq(x1)
     curve(isoq,from=0,to=6,yaxs="i",xaxs="i",ylim=c(0,16),xlab="Quantity of Coal",
           ylab="Quantity of Natural Gas")
     segments(x1,0,x1,x2,lty=2)
     segments(0,x2,x1,x2,lty=2)
rm(a,b,A,Q,isoq,x1,x2)
```

6. ***Copper Mine*** (***): Suppose you are running a copper mine. You can either use machinery or workers to extract the copper. Currently, you are using a 600 horse power train to get the copper out of the mine. You have signed a leasing agreement for that train which you cannot change in the short-run. In the figure below, you will see two isoquant curves and a point indicating the initial (optimal) combination of workers and the train. The input prices remain unchanged throughout the question. Sketch the corresponding isocost line through the initial, optimal point. Assume that the machinery is fixed at 600 horse power and cannot be changed in the short-run. Due to changed demand, you need to expand production to the higher isoquant line. Show the new isocost line? Is it optimal? If yes, why? If no, why not? You are now able to adjust workers and machinery in the long-run while staying on the higher isoquant curve. What is the effect on cost? What is the effect on the optimal input combination. Support your answer in the graph.
```{r PROmining,echo=FALSE}
pK                      = 3
pL                      = 2
a                       = 0.4
Kbar                    = 0.6
isoquantfun             = function(x){(Q/(x^(1-a)))^(1/a)} 
isoquantzerofun         = function(x){(Kbar-(Q/(x^(1-a)))^(1/a))^2} 
isocostfun              = function(C,L){(C-pL*L)/pK}
     Q        = 0.5
     curve(isoquantfun,from=0,to=2,yaxs="i",xaxs="i",ylim=c(0,2),ylab="Machinery (in 1000 hp)",xlab="Workers (in 1000)")
     Lstar1   = optim(par=1,isoquantzerofun,method="Brent",lower=0,upper=1e6)
     points(Lstar1$par,0.6)
     Q        = 0.75
     curve(isoquantfun,from=0,to=2,yaxs="i",xaxs="i",ylim=c(0,2),add=TRUE)     
     Lstar2   = optim(par=1,isoquantzerofun,method="Brent",lower=0,upper=1e6)
rm(pK,pL,a,Kbar,isoquantfun,isoquantzerofun,isocostfun,Q,Lstar1,Lstar2)
```
