# Public Policy
The requirements for markets to allocate resources efficiently are (1) that markets exist for all goods and services in the economy, (2) perfect competition, (3) perfect information, (4) property rights are assigned to all resources and commodities, and (5) there are no externalities. In the following sections, we look at cases where those conditions are violated and where public policy can intervene to improve or restore efficiency. There are video associated with the concepts covered in this chapter:

- [Positive and Negative Externalities](https://youtu.be/bXkJ2aCP4-4)
- [Pollution Control Policies](https://youtu.be/fbDIqlCzD4E)
- [Asymmetric Information and Insurance Markets](https://youtu.be/k3veHT4rSiU)

## Externalities
Externalities are costs (negative externalities) or benefits (positive externalities) on non-market participants. There are no incentives for market participants to take those (social) costs and benefits into account and thus, they base their decisions only on private benefits and costs. Consider the following examples of externalities:

- The manufacturer of cigarettes observes private marginal cost, i.e., the marginal cost of producing cigarettes, but ignores the cost of negative health outcomes that is paid by society. For example, the marginal cost of producing a cigarette may be 3 cents and the marginal external cost of smoking the cigarette may be 2 cents. Thus, the private marginal cost is 3 cents but the social marginal cost is 5 cents. The cigarette manufacturer bases its production decision on 3 cents. A social marginal cost above the private marginal cost is an example of a negative externality. The firm's supply curve does not capture all social marginal costs associated with its product.
- Consider the production of sugary and fatty food and drink items. Producers and consumers of those items do not take the cost into account that the resulting increase in obesity has on non-market participants. The following was published in The Seattle Times  on December 21, 2011 under the title [As passengers add girth, ferries drop capacity](https://www.seattletimes.com/seattle-news/as-passengers-add-girth-ferries-drop-capacity/):

    >The Washington state ferry service isn't going to start turning away hefty passengers, but it has had to reduce the capacity of the nation's largest ferry system because people have been packing on the pounds. Coast Guard vessel-stability rules that took effect nationwide Dec. 1 raised the estimated weight of the average adult passenger to 185 pounds from the previous 160 pounds, based on population information from the Centers for Disease Control and Prevention (CDC). During the past 20 years, there has been a dramatic increase in obesity in the United States and about one-third of American adults are now considered obese, the CDC says on its website. The state ferry system has complied with the new stability rules by simply reducing the listed capacity of its vessels, Coast Guard Lt. Eric Young said Wednesday. "That has effectively reduced the amount of passengers by about 250 passengers or so depending on the particular ferry," said Young, who is based in Seattle. "They generally carry about 2,000, so it's down to 1,750 now." With that many passengers, the ferry wouldn't tip over even if everyone ran to the side at the same time to look at a pod of killer whales, he said.
    
- An example of a positive externality is the upkeep of your yard. If you keep your yard clean and even do some landscaping, you are not only increasing the value of your own home but also the value of the homes around you. For example, your neighbor may be able to more easily sell her or his house at a higher price.

There is a mismatch between social marginal benefits/costs and private marginal benefits/costs which leads to an inefficient allocation of resources. As with the monopoly, there is a deadweight loss from externalities. Externalities arise if property rights have not been established, e.g., clean air, toys for children, fishing in a lake, congestion. People argue whether externalities are a market failure or failure of a markets to exist. For example, a polluting firm is essentially using "clean air" as an input for which it is not paying for. If a market existed for this input, as it does for labor and capital, the input would be priced in the market, i.e. the owner of "clean air" would have to agree to sell just as the owner of labor agrees to sell. Similarly, a positive externality is a product that is produced but no market exists and is consumed for free. Whether we are talking about positive or negative externalities, a deadweight loss exists and a good is either over-produced (negative externality) or under-produced (positive externality). The four figures below demonstrate the inefficiency of a negative externality. 

```{r PUPnegextsetup,echo=FALSE,results=FALSE}
funpmc                  = function(x){10+2*x^1.1}
funsmc                  = function(x){funpmc(x)+10+x^1.1}
fundemand               = function(x){100-4*x}
equimarket              = function(x)(funpmc(x)-fundemand(x))
equiefficicent          = function(x)(funsmc(x)-fundemand(x))
qmarket                 = lsqnonlin(equimarket,1)
qefficient              = (lsqnonlin(equiefficicent,1))
```


```{r PUPnegext1,echo=FALSE,fig.cap="Consumer and Producer Surplus"}
curve(funpmc,from=0,to=25,ylim=c(0,100),yaxs="i",xaxs="i",xlab="Quantity",ylab="Price")
    x       = seq(0,qmarket$x,0.01)
    y1      = funpmc(x) 
    y2      = fundemand(x) 
    polygon(c(x,rev(x)),c(y1,rev(y2)),col="seagreen")
    curve(funsmc,from=0,to=25,ylim=c(0,100),add=TRUE,yaxs="i",xaxs="i")
    curve(fundemand,from=0,to=25,ylim=c(0,100),add=TRUE,yaxs="i",xaxs="i")
    segments(qmarket$x,0,qmarket$x,funsmc(qmarket$x),lty=2)
    segments(qefficient$x,0,qefficient$x,funsmc(qefficient$x),lty=2)
    segments(0,funpmc(qmarket$x),qmarket$x,funpmc(qmarket$x),lty=2)
    segments(0,funpmc(qefficient$x),qefficient$x,funpmc(qefficient$x),lty=2)
```

```{r PUPnegext2,echo=FALSE,fig.cap="Societal Cost"}
curve(funpmc,from=0,to=25,ylim=c(0,100),yaxs="i",xaxs="i",xlab="Quantity",ylab="Price")
    x       = seq(0,qmarket$x,0.01)
    y1      = funpmc(x) 
    y2      = funsmc(x) 
    polygon(c(x,rev(x)),c(y1,rev(y2)),col="red4",border=NA)
    curve(funsmc,from=0,to=25,ylim=c(0,100),add=TRUE,yaxs="i",xaxs="i")
    curve(fundemand,from=0,to=25,ylim=c(0,100),add=TRUE,yaxs="i",xaxs="i")
    segments(qmarket$x,0,qmarket$x,funsmc(qmarket$x),lty=2)
    segments(qefficient$x,0,qefficient$x,funsmc(qefficient$x),lty=2)
    segments(0,funpmc(qmarket$x),qmarket$x,funpmc(qmarket$x),lty=2)
    segments(0,funpmc(qefficient$x),qefficient$x,funpmc(qefficient$x),lty=2)
```

```{r PUPnegext3,echo=FALSE,fig.cap="Deadweight Loss"}
curve(funpmc,from=0,to=25,ylim=c(0,100),yaxs="i",xaxs="i",xlab="Quantity",ylab="Price")
    x       = seq(qefficient$x,qmarket$x,0.01)
    y1      = fundemand(x) 
    y2      = funsmc(x) 
    polygon(c(x,rev(x)),c(y1,rev(y2)),col="darkblue",border=NA)
    curve(funsmc,from=0,to=25,ylim=c(0,100),add=TRUE,yaxs="i",xaxs="i")
    curve(fundemand,from=0,to=25,ylim=c(0,100),add=TRUE,yaxs="i",xaxs="i")
    segments(qmarket$x,0,qmarket$x,funsmc(qmarket$x),lty=2)
    segments(qefficient$x,0,qefficient$x,funsmc(qefficient$x),lty=2)
    segments(0,funpmc(qmarket$x),qmarket$x,funpmc(qmarket$x),lty=2)
    segments(0,funpmc(qefficient$x),qefficient$x,funpmc(qefficient$x),lty=2) 
```

```{r PUPnegext4,echo=FALSE,fig.cap="Efficient Outcome"}
curve(funpmc,from=0,to=25,ylim=c(0,100),yaxs="i",xaxs="i",xlab="Quantity",ylab="Price")
    x       = seq(0,qefficient$x,0.01)
    y1      = funsmc(x) 
    y2      = fundemand(x) 
    polygon(c(x,rev(x)),c(y1,rev(y2)),col="gold")
    curve(funsmc,from=0,to=25,ylim=c(0,100),add=TRUE,yaxs="i",xaxs="i")
    curve(fundemand,from=0,to=25,ylim=c(0,100),add=TRUE,yaxs="i",xaxs="i")
    segments(qmarket$x,0,qmarket$x,funsmc(qmarket$x),lty=2)
    segments(qefficient$x,0,qefficient$x,funsmc(qefficient$x),lty=2)
    segments(0,funpmc(qmarket$x),qmarket$x,funpmc(qmarket$x),lty=2)
    segments(0,funpmc(qefficient$x),qefficient$x,funpmc(qefficient$x),lty=2)
rm(list=ls())    
```
        
### Pigouvian Tax
A Pigouvian tax is a per-unit production tax equal to the difference between social marginal cost and private marginal cost. It is difficult to implement because of unknown social costs of externality. By reducing the market quantity to the efficient quantity, a Pigouvian tax eliminates the DWL from the pollution externality. Difference between Pigouvian tax and pollution tax is that the former is a per-unit tax on output and the latter is a per-unit tax on pollution.

### Coase Theorem
The Coase Theorem states that if transaction costs are low or zero, the establishment of clear property rights will lead to an efficient allocation of resources. Participating parties will negotiate until a mutually advantageous (and efficient) solution is reached. This implies that no government intervention is necessary as long as property rights are established. Note, that the allocation of wealth and income will vary depending on how the property rights are established but that this does not affect efficiency. For example, assume a baker and a doctor are in the same building. The baker needs to run noisy machinery in order to bake bread and make a living. The doctor needs silence in order to practice in his office. The problem in this case is that the property rights with regard to the "noise level" in the building are not clearly established. Now suppose that the doctor could soundproof the walls at a cost of \$100 and the baker could purchase quite machinery at a cost of \$50. If the property "noise level" is given to the doctor, then the baker will purchase quite machinery for \$50. If the noise level is attributed to the baker, the doctor will pay the baker \$50 to purchase new and more quite machinery. The efficient solution, i.e., purchasing machinery for \$50, is reached independently of who is assigned the property rights.

### Pollution Control
What is the optimal level of pollution? Waste is produced in almost any activity and thus, the optimal level of pollution is not equal to zero. Optimal level of pollution is a function of the social cost associated with pollution, e.g., effect on human health, aesthetic effects, as well as the cost of reducing pollution. The cost of reducing pollution is represented by the marginal abatement cost function. We will be looking at three different pollution control policies: 

1. Command-and-control: Under a command-and-control policy, there are either design standards that require the use of a particular technology or performance standards that set the maximum pollution level for individual sources. This policy achieves the desired outcome in terms of pollution but is inefficient.

2. Emission tax: In an emission/pollution tax scenario, firms set the marginal cost of abatement equal to the tax rate and the aggregate cost for a given level of pollution is minimized. An emission tax is efficient.

3. Cap-and-trade: Under a cap-and-trade policy, firms set the marginal cost of abatement equal to the permit price. Problems with cap-and-trade: leakage, e.g., fertilizer, coal, etc. A cap-and-trade policy is efficicent. 

Examples of cap-and trade in the U.S. are the Acid Rain Program and the Clean Air Act. Sulfur dioxide (SO$_2$) and nitrogen oxides (NOx) react in the atmosphere (with water oxygen and oxidants) to form various acid compounds. Health concern: detrimental effects. Ozone affects respiratory illnesses and contributing to permanent lung damage. Market based instruments focus on the quantity and cost of pollution reduction and not who generates the pollution. The [animated maps on the website of the National Atmospheric Deposition Program](https://nadp.slh.wisc.edu/data/animaps.aspx) illustrate the outcomes of those policies using the National Trends Network (NTN) concentration and deposition maps for a variety of pollutants. 

Assume two firms, 1 and 2, with the following total abatement cost function and marginal abatement cost function:
$$TC_1(a_1) = 3 a_1^2 \quad \Rightarrow \quad MC_1(a_1) = 6 a_1 \\
        TC_2(a_2) = 2 a_2^2 \quad \Rightarrow \quad MC_2(a_2) = 4 a_2$$

Suppose that both firms emit 50 tons of pollution per year and the government wants to reduce pollution by 10 tons. Consider the following three scenarios: (1) Proportional reduction, (2) emission tax of \$24 per ton, and (3) cap-and-trade system with a cap set to 40 tons.

###  Policy Examples: SO$_2$ Allowance Trading Program
Theoretically, a well designed pollution tax or cap-and-trade can minimize the total cost of achieving environmental protection [Stavins (1998)](https://doi.org/10.1257/jep.12.3.69). The SO$_2$ Allowance Trading Program established under Title IV of the Clean Air Act amendment of 1990 aimed at reducing emissions by 10 million tons over 1980 levels. The purpose of the program was to control for acid rain to reduce acidification of forest and aquatic ecosystems. However, there have been significant benefits to human health as well . In this program, it was also possible to bank allowances for later use. The compliance with the program was enforced by a penalty of \$2,000 per ton of emissions exceeding the allowance together with the requirement to offset the emissions the following year. The program was mostly successful due to a liberalization in railroads [Schmalensee and Stavins (2013)](https://doi.org/10.1257/jep.27.1.103). Also, the cost of the program was ten times cheaper than initially estimated [Kerr (1998)](https://doi.org/10.1126/science.282.5391.1024).

## Public Goods
Public goods can be consumed by more than one individual at the same time, e.g., national defense, police officers in a city (up to a certain extent). A good is rival if it cannot be consumed by more than one individual at a time. It is excludable if non-paying consumers can be excluded from consumption. Our primary focus so far has been on pure private goods - goods that are both rival and excludable. The Tragedy of the Commons arises when rival goods are made non-excludable through common ownership. Public goods are both non-rival and non-excludable; e.g. national defense, lighthouse. Club goods are non-rival but excludable; e.g. public swimming pools.

|                    |     Rival    |  Non-Rival  |
|--------------------|:------------:|:-----------:|
| **Excludable**     | Private Good |  Club Good  |
| **Non-Excludable** |  Common Good | Public Good |

Common goods lead to the ``Tragedy of the Commons'', e.g., swimming pool as a club good, free rider problem. Suppose that two future roommates need to decided whether to bring a TV to the dormitory. The value of a TV for each player is \$100 and the cost of a TV is \$150.

|              |         | Player B        ||
|:-------------|:----------|:-------:|:------:|
|              |         | Buy     | Dont' buy   |
| **Player A** | Buy    |    1,1  | 20,0   |
|              | Don't buy |   0,20  | 5,5    |

Suppose that the utility of a public good associated with its total quantity $Q_T$ can be represented as follows for consumer can bet written as:
$$U_A(Q_T) = 10 \cdot Q_T^2 = 10 \cdot (Q_A+Q_B)-(Q_A+Q_B)^2$$
And the utility for consumer *B* is expressed as:
$$U_B(Q_T) = 10 \cdot Q_T-\frac{Q_T^2}{2} =10 \cdot (Q_A+Q_B) -\frac{(Q_A+Q_B)^2}{2}$$
The cost of acquiring the good is $C(Q_T)=8(Q_A+Q_B)$. So the total benefit to society can be written as
$$B_S(Q_A,Q_B) = 20 \cdot (Q_A+Q_B)-1.5 \cdot (Q_A+Q_B)^2-8 \cdot (Q_A+Q_B)$$
Solving the first-order conditions leads to $20-3 \cdot (Q_A+Q_B)=8$. Thus, societal benefit is maximized if $Q_T=4$. The key characteristic of a public good that the unit purchased by one consumer can also be consumed by all other consumers. So if $MB_a=10-2q$, $MB_b=10-q$, and $MC=8$, then we have

$$MB_a = MC \quad \Rightarrow \quad q = 1\\
  MB_b = MC \quad \Rightarrow \quad q = 2$$

But since individual $A$ has already purchased 1 unit, individual $B$ will free ride and they will end up with 2 units.

## Natural Resource Management
This section considers two common examples of resource management:

- Fisheries
- Forest Rotation

Before introducing the forestry example, the topic of Net Present Value (NPV) must be presented. Although dynamic aspects of economics are covered later in the book, NPV must be introduced to understand the forestry example. 

### Net Present Value
Net present value (NPV) analysis is based on the idea that a dollar today is different from a dollar tomorrow. Suppose a person has the choice of receiving \$1,000 either today or in a year from now. The person would certainly take the money today and not wait a year because they can take the money know, put it in a bank account, and receive the \$1,000 plus interest in a year from now. For example, assume that the interest rate is 2\% per year. If you make a bank deposit of \$1,000 today ($t=0$), then the value of the bank account in one year will be $V_1 = \$1,000 \cdot 1.02 = \$1,020$. The bank account value in two years ($t=2$) will be $V_2=\$1,000 \cdot 1.02 \cdot 1.02=\$1,040.40$. The general equation (compound interest) for the bank account value is $n$ years given an interest rate of $r$ is:
$$V_n = \$1,000 \cdot (1+r)^n$$

The opposite of compound interest is discounting. Assume a stream of income of \$100 at the end of each year over the next three years.
$$NPV = \frac{100}{1+r} + \frac{100}{(1+r)^2} + \frac{100}{(1+r)^3}$$
Assuming that you receive \$100 at the end of the following three years with an interest rate of 2\%, then the NPV of that stream is \$288.39. Put differently, if you deposit \$283.39 today, you can withdraw exactly \$100 at the end of years 1, 2, and 3. The net present value formula of a stream of $X_t$ per year $t$:
$$NPV = \sum_{t=1}^T \frac{X_t}{(1+r)^t}$$
Net present value calculations are often used in investment decisions by firms but also in court cases relating to wrongful death. In particular, if a person dies while still working, families can seek compensations in the amount of the net present value of earnings. For example, assume that a person who was supposed to retire at age 65 dies in a car accident at age 60. That person was earning \$50,000 at the time of death and the expected growth rate ($g$) of the earnings were 2\% per year. Note that the person could have also died of other causes between the age of 60 and 65 and thus, we need to account for that probability ($d$). To do so, we can consult so called [life tables compiled by the Centers for Disease Control and Prevention (CDC)](https://www.cdc.gov/nchs/products/life_tables.htm). The 2014 table reveal the probabilities of dying displayed in the table below.

| Age  | 60-61    | 61-62    | 62-63    | 63-64    | 64-65    |
|:-----|----------|----------|----------|----------|----------|
|$d_t$ | 0.008959 | 0.009606 | 0.010288 | 0.010995 | 0.011735 |

Hence, if we assume the person would have received \$50,000 this year and over the next five years, we can use the following equation to calculate the net present value of earnings given a discount rate of 5\%.

$$NPV = 50,000 + \sum_{t=1}^5 \frac{50,000 \cdot (1+g)^t \cdot (1-d_t)}{(1+0.05)^t}$$
Hence, the net present value of earnings is \$277,013.36. Note that the perpetual stream of $X$ dollars
$$NPV = \frac{X}{r}$$
This means that the present value of an infinite stream of money is finite.

### Fisheries
Let $g=f(S)$ denote the net growth rate, i.e., births minus deaths excluding harvest, of a fish species. The fish population stock is denoted with $S$ and the maximum fish population is denoted with $k$ and is often referred to as the carrying capacity with unsustainable quantity after this amount. Assume that the function takes the following functional form:
$$g = S \cdot \left(1-\frac{S}{k} \right)$$
This growth function is depicted in the figure below assuming a carrying capacity of $k=100$:

```{r PUBfisheries,echo=FALSE,fig.cap="Growth rate of fish as a function of stock"}
k = 100
g = function(S){S*(1-S/k )}
curve(g,from=0,to=100,xlab="Fish Population Stock (S)",ylab="Growth rate (g)",yaxs="i",xaxs="i",ylim=c(0,30))
```

Before the peak, fish rapidly grow and there are plenty of food and space for the species. After the peak, depletion of the fish by the environment, food scarcity, and less space for species. It can be shown that the maximum sustainable yield is $k/2$. Assuming $k=100$, the maximum amount of fish that can be harvested in a sustainable way from the fishery is 50. 

Next, assume that harvest is a result of the fish population and effort (e.g., number of fishing boats), i.e., $h=f(S,E)$, with $S$ representing fish population and $E$ representing effort. Assume that the harvest function can be written as:
$$h=q \cdot S \cdot E$$
where $q$ represents a coefficient that measures the each of harvesting fish. More effort leads to a higher harvest. Higher fish population also leads to higher harvest. This means that there can be the same harvest rate but different levels of fish population and effort. In the steady state, the harvest rate equals net birth rate of the fish, i.e., $g=h$. To solve for the level of effort at the maximum sustainable yield, the growth and harvest function must be set equal to each other:
$$S \cdot \left(1-\frac{S}{k} \right) = q \cdot S \cdot E$$
Isolating $S$ leads to the following:
$$S=(1-q \cdot E) \cdot k$$
Substituting the fish stock as a function of effort into the harvest function leads to:
$$h= q \cdot (1-q \cdot E) \cdot k \cdot E = q\cdot k \cdot E- k\cdot q^2 \cdot E^2$$
Taking the first derivative and setting it to zero to find the effort that maximizes harvest leads to:
$$q \cdot k -2 \cdot k \cdot q^2 \cdot E = 0$$
Thus, the effort at the maximum sustainable yield is
$$E^{MSY}=\frac{1}{2 \cdot q}$$
Next, the price of fish and the cost of effort must be taken into account. Assume that the price of fish is $p$ and that the cost of effort is $c$. Thus the profit equation can be written as
$$\pi=p \cdot h-c \cdot E$$
Remembering that the condition for sustainability dictates that $g=h$, the harvest $h$ in the above equation can be replaced with the growth function:
$$\pi=p \cdot S \cdot \left(1-\frac{S}{k} \right) -c \cdot E$$
Note that the fish stock is a function of $E$ as well. From the harvest function:
$$S=\frac{h}{q \cdot E}$$ 
And replacing $h$ in the above equations with the function linking harvest to effort:
$$S=\frac{q\cdot k \cdot E- k\cdot q^2 \cdot E^2}{q \cdot E} = k \cdot (1 -q \cdot E)$$
This results in the following profit function:
$$\pi=p \cdot q \cdot k \cdot E - p \cdot k \cdot q^2 \cdot E^2 -c \cdot E$$
Thus, the profit maximizing effort is 
$$E=\frac{p \cdot q \cdot k - c}{p \cdot q^2 \cdot k}$$
This is lower than the maximum sustainable yield. 

### Forest Management
This section covers the optimal harvest of trees, which is also known as rotation period. The growth function of trees is usually assumed to be S-shaped, i.e., slow growth in early and late years but high growth in the intermediate years. Assume a tree growth function of the following function form:
$$g=a_1 \cdot t + a_2 \cdot t^2 + a_3 \cdot t^3$$
where $g$ represents the total timber volume in cubic meters and $t$ represents time. Assume that the growth parameters $a_1$, $a_2$, and $a_3$ take the values 40, 3, and -0.02, respectively. Then over a 100-year period, the total volume of the tree looks as depicted in the figure below.

```{r PUBtreevolume,echo=FALSE}
a1 = 40
a2 =  3
a3 = -0.02
volume = function(t){a1*t+a2*t^2+a3*t^3}
curve(volume,from=0,to=100,xlab="Time",ylab="Volume in Cubic Meters",yaxs="i",xaxs="i",ylim=c(0,15000))
rm(a1,a2,a3,volume)
```

Assume that the price of timber per cubic meter is \$2 and that the harvest cost of the tree consists of a fixed and variable part. The fixed cost is $1,000 and the variable cost is \$0.3 per cubic meter. The net present value is calculated for three different interest rates, i.e., 1\%, 3\%, and 5\%.

```{r PUBforestnpv,echo=FALSE}
a1              = 40
a2              = 3
a3              = -0.02
p               = 2
FC              = 1000
VC              = 0.3
volume          = function(t){a1*t+a2*t^2+a3*t^3}
treedata        = data.frame(age=0:100)
treedata$volume = volume(treedata$age)
treedata$growth = NA
for(t in c(16:100)){
  treedata$growth[t] = treedata$volume[t]/treedata$volume[t-1]-1}
profit1 = data.frame(age=treedata$age,interest="1%",
                     profit=((p-VC)*treedata$volume-FC)/(1+0.01)^treedata$age)
profit3 = data.frame(age=treedata$age,interest="3%",
                     profit=((p-VC)*treedata$volume-FC)/(1+0.03)^treedata$age)
profit5 = data.frame(age=treedata$age,interest="5%",
                     profit=((p-VC)*treedata$volume-FC)/(1+0.05)^treedata$age)
profit  = rbind(profit1,profit3,profit5)

```

## Asymmetric Information
Asymmetric information occurs if two or more parties engage in an transaction and at least one party has more information than the other. This causes high cost customer or low quality suppliers to participate in the market without the other party or parties knowing the cost and/or quality issue. The prime example for asymmetric information is the used car market. The seller has more information about the reliability and quality of the car than the buyer. There are multiple strategies to prevent or reduce asymmetric information. Some examples are:

- **Used Car Market**: Companies like Carfax that track the repair and accident history of cars can help uncover possible issues with a used car.
- **Life Insurance**: An insurance company can require a health exam prior to selling a life insurance policy.
- **Labor Market**: Job interview are designed to reduce the asymmetric information for the employer. The potential job candidate has better information on their ability than the employer. 

Assume that the demand for car insurance is $Q=20-2 \cdot P$. The inverse demand is $P=10-Q/2$. Further, the marginal costs associated with a safe and unsafe drivers are $MC_S=2$ and $MC_u=6$, respectively. A perfectly competitive market without asymmetric information results in welfare maximizing marginal cost pricing. For the safe driver, this leads to:
$$2=10-\frac{Q}{2}$$
Thus, quantity and price for safe drivers are $Q=16$ and $P=2$. Similarly, for the unsafe driver:
$$6=10-\frac{Q}{2}$$
Thus, quantity and price for safe drivers are $Q=8$ and $P=6$. Calculating the consumer surplus from this pricing policy leads to $CS_S=\$ 128$ and $CS_U =\$32$. The total surplus of \$160. If the insurance company cannot determine in which category a driver falls, it has to charge a uniform price. Assuming an equal amount of safe and unsafe drivers, the company sets the price at \$4. It can be shown that this leads to a surplus of \$144, which is lower than the \$160 under no asymmetric information. This insurance problem is also illustrated in the video [Asymmetric Information and Insurance Markets](https://youtu.be/k3veHT4rSiU).

## Exercises
1. ***Negative Production Externality I*** (***): Suppose that the inverse demand function for a particular good can be written as $P=400-5 \cdot Q$ and that private marginal cost $PMC=5 \cdot Q$. The additional external damage per unit produced is $D=2 \cdot Q$. Support the answers to the questions below by using a graph. 
    a. Calculate the market price, quantity, and deadweight loss. 
    b. What are the efficient quantity and price? 
    c. Calculate the per-unit tax that would achieve the efficient outcome.

2. ***Negative Production Externality II*** (***): Demand and supply for a good are written as $Q^D=1000-5 \cdot P$ and $Q^S=2 \cdot P-100$, respectively. The marginal external cost is \$7. Support the answers to the questions below by using a graph. 
    a. Calculate the market price, quantity, and deadweight loss. 
    b. What are the efficient quantity and price? 
    c. Calculate the per-unit tax that would achieve the efficient outcome.

3. ***Polluting Monopolist*** (***): John has a monopoly in the oil refinement market. The oil demand function is $P=80-Q$ and the marginal revenue is $MR(Q)=80-2 \cdot Q$. The private marginal cost is $MC=10$. During the refinement process, air, water, and soil pollution occurs at a constant cost of \$5 per unit of oil. Support the answers to the questions below by using a graph.
    a. What are the profit maximizing price and quantity? 
    b. What are the efficient price and quantity? 
    c. Calculate the deadweight loss associated with the monopoly situation? Should the government tax emissions? If yes, at what rate? If no, why?

4. ***Pollination*** (***): Pollination by bees is very important for plant reproduction and substantial fees are paid to beekeepers [(Rucker et al., 2012)](https://doi.org/10.1093/ajae/aas031). Imagine a beekeeper and an apple orchard farmer being neighbors. Note this is a situation of a positive externality. The beekeeper receives the revenue from selling honey but in the absence of any payments, does not receive any money for the bees pollinating nearby orchards or fields. Suppose that one beehive ($H$) can pollinate one hectare. The pollination of an hectare without the bees costs \$20. The beekeeper can sell the honey from a beehive at \$50. The total cost of the beekeeper is $TC=H^2+20$ and marginal costs $MC=2 \cdot H$. Support the answers to the questions below by using a graph.
    a. How many hives would the beekeeper maintain if operating independently of the farmer?
    b. What is the socially efficient number of hives?
    c. In the absence of transaction costs, what outcomes do you expect to arise from bargaining between the beekeeper and the farmer?
    d. How high would total transaction costs have to be to erase all gains from bargaining?

5. ***Efficient Polluting Monopolist*** (**): Externalities and monopoly power lead to a deadweight loss when looked at separately. Using a graph, illustrate the case where a polluting monopolist can be efficient in the absence of any intervention. Draw a linear demand function and the corresponding marginal revenue function. Next, draw an upward sloping marginal cost function starting at the origin (note that the result does not change if you draw the marginal cost function with an intercept). Determine the profit maximizing output and price. Next draw an external marginal cost function such that the initially determine quantity and output are efficient, i.e., with no deadweight loss.

6. ***Pareto vs. Kaldor-Hicks Efficiency*** (**): Consider the following table. Does this project meet the criteria of Pareto efficiency? Kaldor-Hicks efficiency? Would it pass a majority vote?

    | Individual Resident | Individual Benefit | Cost Share   |
    |:--------------------|-------------------:|-------------:|
    | A                   | \$4,500            | \$3,000      |
    | B                   | \$1,000            | \$3,000      |
    | C                   | \$2,500            | \$3,000      |
    | D                   | \$4,500            | \$3,000      |
    | E                   | \$1,500            | \$3,000      |
    |Total Benefit        | \$14,000           | \$15,000     |

7. ***Streetlight Installation*** (**): Three neighbors vote on installing a streetlight. If they vote in favor of the streetlight, each neighbor has to pay \$100. They each value the streetlight at \$200, \$150, and \$50, respectively. Does this project meet the criteria of Pareto efficiency? Kaldo-Hicks efficiency? Would it pass a majority vote?