# Risk and Uncertainty
So far, we considered decisions based on the assumption that there is no uncertainty. In reality, we are surrounded by uncertainty such as changes in prices, income, and so on. Good (or bad) stuff happens. Some examples of risks that affect you are retirement funds (e.g., mutual fund, stock market, etc.), gasoline (energy) prices, accidents (e.g., home, car, health, etc.), or sub-prime mortgages. Most decisions are based on expectations about the future. Many economic questions and issues are associated with risk: How do people make decisions under uncertainty? Is there a market (and hence, a price) for risk? Can we reduce risk?

## Uncertainty
Suppose that you can engage in one of the following three games (or gambles):

- Receiving \$100.
- Flipping a coin and receiving \$200 if the result is heads and \$0 if the result is tails.
- Die roll and receiving \$400 for 1, \$70 for 2, \$55 for 3, \$40 for 4, \$25 for 5, and \$10 for 6.

The expected outcome is \$100 in each case but you certainly have a preference for which option you pick. However, the variances are 0, 10,000, and 18,375. We will soon see that uncertain outcomes are worth less in terms of utility than certain outcomes (like receiving \$100) even when the expected payoffs are identical. A prime example to illustrate this concept is called the St. Petersburg Paradox. The game is simple: Flip a coin and I will pay you \$2$^n$ where $n$ is the number of flips until heads. So the expected payoff is
$$\text{Expected Payoff} = \frac{1}{2} \cdot 2 + \frac{1}{4} \cdot 4 + \frac{1}{8} \cdot 8 + \cdots = \infty$$
An actuarially fair game is where the cost of entry equals the expected payoff. That is $E(X) = \mathrm{Pr}_{win} \cdot \pi_{win} + \mathrm{Pr}_{lose} \cdot \pi_{lose} = 0$.

A probability distribution refers to a list of outcomes and probabilities associated with those outcomes. The sum of the probabilities across all outcomes is always 1. Consider a project investment as outlined in the table below. The profits from projects *A* and *B* depend on the state of the economy. 

| Economic State | Probability | Profit A | Profit B |
|:---------------|:-----------:|---------:|---------:|
| Recession      | 40\%        | 4000     | 0        |
| Mormal         | 35\%        | 5000     | 5000     |
| Boom           | 25\%        | 1000     | 8000     |

To achieve some quantification of risk associated with those projects, we need to calculate the expected value and the variance. The expected value is calculated as follows:
$$E(\pi)  = \sum_i \pi_i \cdot p_i$$
and thus, the expected values of projects A and B are
$$E(\pi_A) =  0.4 \cdot 4000 + 0.35 \cdot 5000 + 0.25 \cdot 1000 = 3600$$
$$E(\pi_A) =  0.4 \cdot 0 + 0.35 \cdot 5000 + 0.25 \cdot 8000 = 3750$$
The variance is calculated as:
$$Var(\pi) = \sum \left[ \pi_i - E(\pi) \right]^2 \cdot p_i$$ 
For projects A and B, we have
$$Var(\pi_A) = 2,440,000       \Leftrightarrow \sigma_A = 1560$$
$$Var(\pi_A) = 10,687,500      \Leftrightarrow \sigma_B = 3269$$


## Expected Utility
The expected utility is the sum of the utilities associated with all possible outcomes, weighted by the probability that each outcome will occur. From Jensen's inequality, we have

- $E[U(w)] = U[E(w)] \Rightarrow$ Risk-neutral
- $E[U(w)] < U[E(w)] \Rightarrow$ Risk-averse
- $E[U(w)] > U[E(w)] \Rightarrow$ Risk-seeking

The risk premium is the maximum amount of money a risk averse person will pay to avoid taking a risk. Note that this is an extremely important concept. For example, the risk premium leads to higher interest rates for consumers or companies with a low credit rating. To illustrate the concept of expected utility, assume that we have a risk averse individual and two possible outcomes: (1) Getting \$1 with a probability of 60\%, and (2) Getting \$8 with a probability of 40\%. The expected value of wealth is
$$E(w)=0.6 \cdot 1+0.4 \cdot 8=3.80$$
And the expected utility is:
$$E(U) = 0.6 \cdot U(\$1) + 0.4 \cdot U(\$8)$$
We also have $u(\$3.8)$. Note that $u(\$3.8) > 0.6 \cdot u(\$1) + 0.4 \cdot u(\$8)$

```{r RSKexpectedutility}
payoff    = c(4,25)
prob      = 0.5
utility   = function(w){w^0.4}
epayoff   = prob*payoff[1]+(1-prob)*payoff[2]
eutility  = prob*utility(payoff[1])+(1-prob)*utility(payoff[2])
     curve(utility,from=0,to=30,yaxs="i",xaxs="i",ylab="Wealth",xlab="Utility")
     segments(payoff[1],0,payoff[1],utility(payoff[1]),lty=2,col="black")
     segments(payoff[2],0,payoff[2],utility(payoff[2]),lty=2,col="black")
     segments(payoff[1],utility(payoff[1]),payoff[2],utility(payoff[2]),lty=2,col="black")
     segments(epayoff,0,epayoff,eutility,lty=2,col="black")
```

## Risk Diversification
The colloquial expression for risk diversification is not to put all eggs in one basket. There are gains from diversification as long as assets are not perfectly correlated. You can think about the stock market or if you have a football team with only one good quarterback not only poor backup quarterbacks. Consider an investment possibility in two firms: raincoats and sunglasses.

- Raincoat company: $p^R_0=10$ and $p^R_1=5$ if sunny or $p^R_1=20$ if rainy
- Sunglasses company: $p^S_0=10$ and $p^S_1=20$ if sunny or $p^S_1=10$ if rainy

In this case, a diversification strategy that allocates \$5 to the raincoat company and \$5 to the sunglasses company allows a return of 25\% if sunny and 50\% if rainy. 

## Insurance Market
If people are risk averse then why do insurance companies exist? The key here is that insurance companies can engage in what is called risk pooling. To illustrate the concept, suppose that you are flipping a coin and must predict the share of tails. The expected value of the share of heads (or tails) is $E(H)=E(T)=0.5$. The variance of $n$ coin tosses is written as
$$Var(n) = \frac{p \cdot (1-p)}{n} = \frac{0.5 \cdot 0.5}{n}$$
Thus, for 1, 10, and 1000 coin flips, the variances are $Var(1)=0.5$, $Var(10)=0.025$, $Var(1000) = 0.00025$, respectively. That is, the variance associated with the share of coin flips decreases with an increasing number of flips. Think about how difficult it is to predict the correct share of heads from a single coin flip but how easy it is to predict the share of heads from several thousand coin flips. The same concept applies for an insurance company. Given a large number of policy holders, there is almost perfect predictability of the number (or share) of claims in a year. Note that the key assumption here is independence of events/claims. 

To illustrate this concept, assume that the probability of a house fire is 1\%. To simulate the number of claims an insurance policy faces each year, a simulation is conducted involving the following steps: 

1. Simulate the damage of $n$ homeowners
2. Calculate the share
3. Repeat 1,000 times
4. Generate histogram

An actuarially fair insurance premium is equal to the probability of a claim times to payout.

```{r}
fireinsurance  = function(ninsur){
     fire = matrix(0,ndraws,1)
     for (i in 1:ndraws){
          fg      = sample(x=c(1,0),size=ninsur,replace=TRUE,prob=c(1/100,99/100))
          fg      = sum(fg)/ninsur
          fire[i] = fg}
     return(fire)}
ndraws = 1000
fire1  = fireinsurance(1000)
fire2  = fireinsurance(10000)
fire3  = fireinsurance(25000)
fire4  = fireinsurance(100000)
par(mfrow=c(2,2))
  hist(fire1,xlim=c(0,0.025),ylim=c(0,750),main="1000 People Insured",xlab="",
       breaks=seq(0,0.03,0.001))
  hist(fire2,xlim=c(0,0.025),ylim=c(0,750),main="10000 People Insured",xlab="",
       breaks=seq(0,0.03,0.001))
  hist(fire3,xlim=c(0,0.025),ylim=c(0,750),main="25000 People Insured",xlab="",
       breaks=seq(0,0.03,0.001))
  hist(fire4,xlim=c(0,0.025),ylim=c(0,750),main="100000 People Insured",xlab="",
       breaks=seq(0,0.03,0.001))
rm(fireinsurance,ndraws,fire1,fire2,fire3,fire4)
```

## Exercises

1. ***Expected Utility*** (**): The utility function of an individual is written as $U=\sqrt{M}$. The person faces a gamble in which income is \$36 with probability 1/2 and \$100 with probability 1/2. 
    a. What payoff with certainty is indifferent to this gamble? 
    b. What is the risk premium associated with this gamble? 
    c. How much would the individual be willing to pay to avoid taking the risk?